High-Fidelity Counterfactual Explanations through Conformal Prediction

### 2023-08-10 dev/rebuttal/support.qmd

---
format: 
    pdf:
        fontsize: 10pt
---

![Turning a 6 into a 0. Top row: ECCCo. Bottom row: Wachter. The different underlying models across columns: (a) MLP, (b) Small Ensemble $n=5$, (c) Large Ensemble $n=50$, (d) LeNet-5, (e) JEM, (f) JEM Ensemble.](www/mnist_6to0_7.png){#fig-mnist-1 width="80%"}

![Turning a 9 into a 7. Rows and columns as in @fig-mnist-1.](www/mnist_9to7_1.png){#fig-mnist-2 width="80%"}

<!-- ![Turning a 4 into a 1. Top row: ECCCo. Bottom row: Wachter. The different underlying models across columns: (a) MLP, (b) Small Ensemble $n=5$, (c) Large Ensemble $n=50$, (d) LeNet-5, (e) JEM, (f) JEM Ensemble.](www/mnist_4to1_11.png){#fig-mnist-3 width="75%"} -->

>>>>

---
format: 
    pdf:
        fontsize: 10pt
        geometry: 
            - top=10mm
            - bottom=10mm
            - left=10mm
            - right=10mm
---

\pagenumbering{gobble}

::: {#fig-fmnist layout="[[-1,10,-1], [-1,10,-1], [-1,10,-1], [20,-1,20,-1,20]]"}

![Turning a 9 into a 7.](www/mnist_9to7_1.png){#fig-mnist-1 width="50%"}

![Turning a 2 into a 3. ](www/mnist_2to3_14.png){#fig-mnist-2 width="50%"}

![Turning a 2 into a 3.](www/mnist_4to1_12.png){#fig-mnist-3 width="50%"}

![Turning a dress (left) into a trouser (right): a gap between the legs appears as expected.](www/fmnist_dress.png){#fig-fmnist-dress width="5cm"}

![Turning a pullover (left) into a coat (right): it looks like a V-neck has formed.](www/fmnist_pullover.png){#fig-fmnist-pullover width="5cm"}

![Turning a boot (left) into a sandal (right): pixels are darkened in the right places.](www/fmnist_boot.png){#fig-fmnist-boot width="5cm"}

Qualitative examples for MNIST and Fashion-MNIST. **MNIST** (@fig-mnist-1 to @fig-mnist-3): Top row: ECCCo. Bottom row: Wachter. The different underlying models across columns: (a) MLP, (b) Small Ensemble $n=5$, (c) Large Ensemble $n=50$, (d) LeNet-5, (e) JEM, (f) JEM Ensemble. **Fashion-MNIST** (@fig-fmnist-dress to @fig-fmnist-boot): The underlying classifier is a small ensemble of 5 MLPs with one hidden layer. Counterfactuals are generated by ECCCo. 
:::

### 2023-08-10 notebooks/fmnist.qmd

```{julia}
function plot_fmnist(
    x::Union{AbstractArray, Int}=x_factual, target::Int=target;
    generator = ECCCoGenerator(),
    rng::Union{Int,AbstractRNG}=Random.GLOBAL_RNG,
    T::Int = 100,
    use_class_loss::Bool = true,
    model=model_dict["Ensemble"],
    img_height::Int = img_height,
    test_data::Bool = false,
    kwrgs...,
)

    # Setup:
    Random.seed!(rng)
    if x isa Int
        x_fact = counterfactual_data.X[:,rand(findall(labels.==x))][:,:]
    else
        x_fact = x
    end

@@

    ce = generate_counterfactual(
        x_fact, target, data, model, generator; 
        decision_threshold=γ, max_iter=T,
        initialization=:identity,
        converge_when=:generator_conditions,
    )

    # Plot:
    p1 = Plots.plot(
        convert2image(MNIST, reshape(x_fact,28,28)),
        axis=([], false),
        size=(img_height, img_height),
        title="Factual"
    )

    _x = CounterfactualExplanations.counterfactual(ce)
    p2 = Plots.plot(
        convert2image(MNIST, reshape(_x,28,28)),
        axis=([], false), 
        size=(img_height, img_height),
        title="Counterfactual"
    )

    plt = Plots.plot(p1, p2; size=(img_height*2,img_height), layout=(1,2), kwrgs...)

@@

```{julia}
rng = rand(1:10000)
factual = 3
plt = plot_fmnist(3, 1, rng=rng)
```

>>>>

```{julia}
label_names = [
    "T-shirt/top",
    "Trouser",
    "Pullover",
    "Dress",
    "Coat",
    "Sandal",
    "Shirt",
    "Sneaker",
    "Bag",
    "Boot",
]

function plot_fmnist(
    x::Int=3, targets::Vector{Int}=[1];
    generator = ECCCoGenerator(),
    rng::Union{Int,AbstractRNG}=Random.GLOBAL_RNG,
    T::Int = 100,
    use_class_loss::Bool = true,
    model=model_dict["Ensemble"],
    img_height::Int = img_height,
    test_data::Bool = false,
    init::Symbol = :identity,
    noise::Float32 = 0.0f0,
    kwrgs...,
)

    # Setup:
    Random.seed!(rng)
    x_fact = counterfactual_data.X[:,rand(findall(labels.==x))][:,:]
    x_fact = pre_process(x_fact, noise=noise)

@@

    # Plot:
    p1 = Plots.plot(
        convert2image(MNIST, reshape(x_fact,28,28)),
        axis=([], false),
        size=(img_height, img_height),
        title="$(label_names[x+1]) (factual)",
        dpi=300
    )

    plts = [p1]

    for t in targets
        ce = generate_counterfactual(
            x_fact, t, data, model, generator; 
            decision_threshold=γ, max_iter=T,
            initialization=init,
            converge_when=:generator_conditions,
        )
        _x = CounterfactualExplanations.counterfactual(ce)
        plt = Plots.plot(
            convert2image(MNIST, reshape(_x,28,28)),
            axis=([], false), 
            size=(img_height, img_height),
            title="$(label_names[t+1])", 
            dpi=300
        )
        plts = [plts..., plt]
    end

    plt = Plots.plot(plts..., size=(img_height*length(plts),img_height), layout=(1,length(plts)), kwrgs...)

@@

```{julia}
rng = 42
gen = ECCCoGenerator(
    λ=[0.1,0.30,0.30], 
    nsamples=25,
    nmin=10,
)
plts = []

plt = plot_fmnist(3, [1]; rng=rng, generator=gen, img_height=200)
display(plt)
push!(plts, plt)
savefig(plt, "dev/rebuttal/www/fmnist_dress.png")

plt = plot_fmnist(9, [5]; rng=rng, generator=gen, img_height=200)
display(plt)
push!(plts, plt)
savefig(plt, "dev/rebuttal/www/fmnist_boot.png")

plt = plot_fmnist(2, [4]; rng=rng, generator=gen, img_height=200)
display(plt)
push!(plts, plt)
savefig(plt, "dev/rebuttal/www/fmnist_pullover.png")
```
